{\huge Changes and trends in budburst and leaf flush across Europe and North America} \\A meta-analysis of local adaptation in spring phenology studies

### 2023-04-20 Docs/manuscript.Rnw
Some notes on results

\section{Introduction}
\begin{itemize}
\item Define spring event, fall event
\item Background
\item Importance: An accurate understanding of how species distribution might be influenced by environmental conditions at a continental scale is critical for future range shift prediction. 
\item Knowledge Gap: Anecdotally, the spring phenology of trees from European provenances appeared to exhibit stronger clines and less plasticity than in North American studies. Despite a growing interest in predicting local adaptation across locations, no study has comprehensively examined clines for spring and fall events and what factors may underlie differences observed across studies.
\item Research question (links to results)
\end{itemize}

@@

\section{Methods}
\subsection{Data Collection}
%emw2Mar: Try to open with WHY (goal/aim) you did something more than what you did, not a rule, but try to do when possible. 
To locate common garden studies that documented the timing of spring events of woody plant species we searched and reviewed the peer-reviewed literature. On 14 December 2022 we searched Web of Science (Thompson Reuters, New York, NY) using the following terms:
\begin{quote}
TOPIC = (common garden* OR provenance*) AND (leafout* OR leaf out* OR budburst OR spring phenolog*)
\end{quote}
which returned 122 publications. We also contacted authors of previous review papers \citep{AitkenBemmels16, Alberto13}, to help further search the literature. We then reviewed the methods and results of all publications to refine to only studies that met the following criteria: (a) focused on woody plants originating from either Europe or North America, (b) had provenance trials/common gardens on the same continent, (c) reported spring events in units of calendar days (day of year or DOY) or could be converted into DOY (see `Additional Methods' section in the Supplement), and (d) reported latitude and longitude of provenances and gardens. 
\\
\\
Based on these criteria we found 18 provenance/common garden studies, distributed throughout North America and Europe, with the majority of data concentrated in North America (Figure X). From each paper we extracted phenological data on spring events (budburst and leaf flush in DOY) and, when present in the same paper, fall events (bud set, leaf senescence, growth cessation, and leaf abscission) by species, provenance and garden location (geographic coordinates). We used ImageJ (version 1.53k; \citealp{schneider_rasband_eliceiri_2012}) to extract values from figures whenever necessary. For studies that reported event dates relative to a reference date other than 1 January (e.g. \citealp{Rehfeldt1994}), we converted such dates to DOY using the Lubridate Package in R.
\improvement{display a table and a map here, and in the caption, clarify that some studies had more than one gardens and two studies shared the same garden}
%emw2Mar -- mv the following sentence to the caption of the map figure: The distribution was skewed toward North America (n = X), with only X located in Europe. %alinaMar5:think abt displaying number of studies or provenances
\\
\\
To understand how climatic differences, in addition to geographical differences, shape local adaptation in spring events we extracted several types of climate data using information about provenance latitude, longitude, and elevation from original publications. We estimated the mean annual temperature (MAT) for each provenance using the Climate Information Tool by Food and Agriculture Organization of the United Nations \citep{FAO2022}and ClimateWNA \citep{wang2016}. To examine more explicitly climate near spring events, we used gridded daily temperature data for 2011 to 2020 for all European and North American provenances and gardens from E-OBS and the daymetr R package respectively \citep{cornes2018,hufkens2018}. Using these data we estimated how much the daily temperatures overlapped between garden and provenance locations, which we call `climate overlap.' For this we used the Overlap package in R to calculate the percentage overlap of the daily temperature of each provenance in spring months (March to May) from 2011 to 2020 in relation to the daily temperature of their corresponding gardens. Finally, using the daily data, we also calculated Growing degree days (GDD), a commonly used heat accumulation measure to forecast phenological development in plants (Miller et al., 2001), based on the accumulation of mean daily temperatures ($T_{m}$)from 2011 to 2022 above a baseline of 0$^{\circ}$C, from January 1 until budburst and leaf flush with the following formula: GDD = $\sum$($T_{m}$ - $0^{\circ}$C for $T_{m}$ $\ge$ $0^{\circ}$C; 0 for $T_{m}$ $\le$ $0^{\circ}$C).

\subsection{Analyses}
To estimate clines in spring and fall phenological events across species we used hierarchical Bayesian models. We regressed day of year (DOY) of events against geographical and climate predictors with partial pooling (sometimes called `random effects') on the intercept and slope for each species within each garden. Because most tree species were present in only one common garden in our dataset, it was impossible to fit garden and species separately, thus we treat each species within a garden as a unique group. Using posterior estimates for each species within a garden, we estimated effects of continent (North America vs. Europe) and species type (angiosperm vs. gymnosperm). All models were fit in rstanarm package (version 2.21.3; \citealp{brilleman2018}), using default priors, with 4 chains and 1000 sampling iterations per chain for a total of 4000 samples. We checked for model fit by ... [ADD here]. We present estimates as mean $\pm$ 90\% uncertainty intervals given parenthetically, unless otherwise stated. 
%emw2Mar -- It would be good if you have tidy code you can post with the data so folks can fit the models themselves. 
%emw2Mar -- Did you use default priors? If so, can you figure out how to increase them threefold and see if answers change (they change run to run, but should not change dramatically)? Ideally they don't and then we can add 'increasing priors three-fold did not change results.'

\section{Results}
%emw24Jan -- below is great opening!
%emw6Mar -- results are in past tense generally
Our final dataset captured seven angiosperm and eight gymnosperm species from 19 common gardens, encompassing 384 North American provenances and 101 European provenances, with observations from 1962 to 2019. Seven species also had fall event information available. 
\\

@@

Overall, spring events were not related to provenance latitude (\Sexpr{format(round(mean(northamerican_springpost_fitC_spring_lat),digits = 2), nsmall=2)} days/degree [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.05), digits = 1)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.95), digits = 2)}]) or MAT (\Sexpr{format(mean(northamerican_springpost_fitC_spring_mat),digits = 1)} days/degree [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.95), digits = 2)}]) in North America. \todo{add figure here} Spring events advanced slightly with provenance latitude (\Sexpr{format(round(mean(european_springpost_fitC_spring_lat),digits = 2),nsmall=2)} days/degree [\Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.95), digits = 2)}]) and MAT (\Sexpr{format(round(mean(european_springpost_fitC_spring_mat),digits = 2),nsmall=2)} days/degree [\Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.95), digits = 2)}]) in Europe, meaning spring events are slightly earlier where provenance MAT is lower (i.e., higher, more northern latitudes)(Fig. \ref{figure:springfall_latmat} \& Supplement Table. \ref{supplement-table:model_sf_lat}), though uncertainty intervals strongly overlapped zero. Results were similar using other distance metrics in lieu of latitude (see Supplement Fig.\ref{supplement-figure:lat_distance} for results using the absolute value of the distance and the spherical distance difference between provenance and garden latitude).

\begin{figure}[!h] 
    \centering
 \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/springfall_latmat.png}
    \caption{Spring event Day of Year (DOY) in relation to provenance latitude and MAT, coded by symbol for species and color for garden with linear fits from hierarchical Bayesian models. Spring events shown on top and fall event at the bottom.} 
    \label{figure:springfall_latmat}
\end{figure}

@@

In contrast, fall events (budset, leaf senescence, leaf abcission) advanced strongly with provenance latitude and MAT, meaning fall events are earlier where provenance MAT is lower. This relationship, however, was observed mostly in North America where fall events advance \Sexpr{format(round(abs(mean(northamericanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(northamericanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT$^{\circ}$C (Fig. \ref{figure:continent}). In Europe, such relationship is weak: advance \Sexpr{format(round(abs(mean(europeanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(europeanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT $^{\circ}$C(Fig. \ref{figure:continent}).

\begin{figure}[!h] 
    \centering
 \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/continent_histogram.png}
    \caption{Caption placeholder}
    \label{figure:continent}
\end{figure}

Effects of provenance latitude on both spring and fall events were similar across angiosperms and gymnosperms. However, effects of MAT on spring events weakly diverge: spring events get earlier as MAT increases in angiosperms and delay as MAT increases in gymnosperms, except for \emph{Pseudotsuga menziesii}. Fall events delay in warmer locations for both species types, but slightly more so for gymnosperms (\Sexpr{format(round(mean(deciduouspost_fitC_fall_mat),digits = 2),nsmall=2)} days VS. \Sexpr{format(round(mean(evergreenspost_fitC_fall_mat),digits = 2),nsmall=2)} days)(Fig. \ref{figure:spp_type}).

@@

We also tested whether climate effects on events may be stronger using metrics from seasonal windows closer to event dates, but results were not qualitatively different than using MAT. We observed very weak effects of climate overlap on spring events, nearly identical across angiosperms and gymnosperms. Fall events diverge as climate overlap declines for both angiosperms and gymnosperms, but slightly more strongly for gymnosperms (Fig.\ref{figure:overlap}).

@@

\begin{figure}[!h] 
    \centering
 \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/climate_overlap.png}
    \caption{Caption placeholder}
    \label{figure:overlap}
\end{figure}
%emw6Mar -- Climate overlap figures can go in supp.

>>>>

%EMWApr20 -- please organize your first outline using \enumerate, which is much easier for me to follow and refer to #s then
\section{Introduction}
\begin{itemize}
\item Define spring event, fall event
\item Background
\item Importance: An accurate understanding of how species distribution might be influenced by environmental conditions at a continental scale is critical for future range shift prediction. 
\item Knowledge Gap: Anecdotally, the spring phenology of trees from European provenances appeared to exhibit stronger clines and less plasticity than in North American studies. Despite a growing interest in predicting local adaptation across locations, no study has comprehensively examined clines for spring and fall events and what factors may underlie differences observed across studies.
\item Research question (links to results)
\end{itemize}

@@

\section{Methods}
\subsection{Data Collection}
%emw2Mar: Try to open with WHY (goal/aim) you did something more than what you did, not a rule, but try to do when possible. 
To locate common garden studies that reported the timing of spring events of woody plant species we searched and reviewed the peer-reviewed literature. On 14 December 2022 we searched Web of Science (Thompson Reuters, New York, NY) using the following terms:
\begin{quote}
TOPIC = (common garden* OR provenance*) AND (leafout* OR leaf out* OR budburst OR spring phenolog*)
\end{quote}
which returned 122 publications. We also contacted authors of previous review papers \citep{AitkenBemmels16, Alberto13}, to help further search the literature. We then reviewed the methods and results of all publications to refine to only studies that met the following criteria: (a) focused on woody plants originating from either Europe or North America, (b) had provenance trials/common gardens on the same continent, (c) reported spring events in units of calendar days (day of year or DOY) or could be converted into DOY (see `Additional Methods' section in the Supplement), and (d) reported latitude and longitude of provenances and gardens. 
\\
\\
Based on these criteria we found 18 provenance/common garden studies, distributed throughout North America and Europe, with the majority of data concentrated in North America (Figure X). From each paper we extracted phenological data on spring events (budburst and leaf flush in DOY) and, when present in the same paper, fall events (bud set, leaf senescence, growth cessation, and leaf abscission) by species, provenance and garden location (geographic coordinates). We used ImageJ (version 1.53k; \citealp{schneider_rasband_eliceiri_2012}) to extract values from figures whenever necessary. For studies that reported event dates relative to a reference date other than 1 January (e.g. \citealp{Rehfeldt1994}), we converted such dates to DOY using the Lubridate Package in R.
\improvement{display a table and a map here, and in the caption, clarify that some studies had more than one gardens and two studies shared the same garden}
%emw2Mar -- mv the following sentence to the caption of the map figure: The distribution was skewed toward North America (n = X), with only X located in Europe. %alinaMar5:think abt displaying number of studies or provenances
%EMWApr20 -- table in supp and map in main text (figure caption should refer to table in supp)
\\
\\
To understand how climatic differences, in addition to geographical differences, shape local adaptation in spring events we extracted several types of climate data using information about provenance latitude, longitude, and elevation from original publications. We estimated the mean annual temperature (MAT) for each provenance using the Climate Information Tool by Food and Agriculture Organization of the United Nations \citep{FAO2022}and ClimateWNA \citep{wang2016}. To examine more explicitly climate near spring events, we used gridded daily temperature data for 2011 to 2020 for all European and North American provenances and gardens from E-OBS and the daymetr R package respectively \citep{cornes2018,hufkens2018}. Using these data we estimated how much the daily temperatures overlapped between garden and provenance locations, which we call `climate overlap.' For this we used the Overlap package in R to calculate the percentage overlap of the daily temperature of each provenance in spring months (March to May) from 2011 to 2020 in relation to the daily temperature of their corresponding gardens. Finally, using the daily data, we also calculated Growing degree days (GDD), a commonly used heat accumulation measure to forecast phenological development in plants (Miller et al., 2001), based on the accumulation of mean daily temperatures ($T_{m}$)from 2011 to 2022 above a baseline of 0$^{\circ}$C, from January 1 until budburst and leaf flush with the following formula: GDD = $\sum$($T_{m}$ - $0^{\circ}$C for $T_{m}$ $\ge$ $0^{\circ}$C; 0 for $T_{m}$ $\le$ $0^{\circ}$C).

\subsection{Analyses}
To estimate clines in spring and fall phenological events across species we used Bayesian hierarchical models. We regressed day of year (DOY) of events against geographical and climate predictors with partial pooling (sometimes called `random effects') on the intercept and slope for each species within each garden. Because most tree species were present in only one common garden in our dataset, it was impossible to fit garden and species separately, thus we treat each species within a garden as a unique group. Using posterior estimates for each species within a garden, we estimated effects of continent (North America vs. Europe) and species type (angiosperm vs. gymnosperm). All models were fit in rstanarm package (version 2.21.3; \citealp{brilleman2018}), using default priors, with 4 chains and 1000 sampling iterations per chain for a total of 4000 samples. We checked for model fit by ... [ADD here]. We present estimates as mean $\pm$ 90\% uncertainty intervals given parenthetically, unless otherwise stated. 
%emw2Mar -- It would be good if you have tidy code you can post with the data so folks can fit the models themselves. 
%emw2Mar -- Did you use default priors? If so, can you figure out how to increase them threefold and see if answers change (they change run to run, but should not change dramatically)? Ideally they don't and then we can add 'increasing priors three-fold did not change results.'

\section{Results}
%emw24Jan -- below is great opening!
%emw6Mar -- results are in past tense generally
Our final dataset included seven angiosperm and eight gymnosperm species from 19 common gardens, encompassing 384 North American provenances and 101 European provenances, with observations from 1962 to 2019. Seven species also had fall event information available. 
\\

@@

%EMWApr20 -- if you look at the intervals the results in NA and Europe seem similar so I changed the text. Also -- you should report the OVERALL model estimate across both europe and NA ...
Overall, spring events were not related to provenance latitude or MAT, neither across continents (Fig. \ref{figure:springfall_latmat} \& Supplement Table. \ref{supplement-table:model_sf_lat}, ADD latitude: stats, MAT: stats), nor in North America (latitude: \Sexpr{format(round(mean(northamerican_springpost_fitC_spring_lat),digits = 2), nsmall=2)} days/degree [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.05), digits = 1)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT \Sexpr{format(mean(northamerican_springpost_fitC_spring_mat),digits = 1)} days/degree [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.95), digits = 2)}]) or Europe (latitude: \Sexpr{format(round(mean(european_springpost_fitC_spring_lat),digits = 2),nsmall=2)} days/degree [\Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT:\Sexpr{format(round(mean(european_springpost_fitC_spring_mat),digits = 2),nsmall=2)} days/degree [\Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.95), digits = 2)}]). Results were similar using other distance metrics in lieu of latitude (see Supplement Fig.\ref{supplement-figure:lat_distance} for results using the absolute value of the distance and the spherical distance difference between provenance and garden latitude). \\
\todo{add figure here} 
% Spring events advanced slightly with provenance latitude in Europe spring events are slightly earlier where provenance MAT is lower (i.e., higher, more northern latitudes), though uncertainty intervals strongly overlapped zero.

%EMWApr20 -- reminder that Fig 1 needs a way to tell apart Europe and NA garden -- you can say in caption for now how the GardenID relates and eventually show on figure (add some text and a bracket to gardenID legend I think)
\begin{figure}[!h] 
    \centering
 \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/springfall_latmat.png}
    \caption{Spring event Day of Year (DOY) in relation to provenance latitude and MAT, coded by symbol for species and color for garden with linear fits from hierarchical Bayesian models. Spring events shown on top and fall event at the bottom.} 
    \label{figure:springfall_latmat}
\end{figure}

@@

%EMWApr20 -- add stats for model (across Europe and NA) in first sentence. 
In contrast, fall events (budset, leaf senescence, leaf abcission) advanced strongly with provenance latitude and MAT, meaning fall events are earlier where provenance MAT is lower. This relationship, however, was observed mostly in North America where fall events advance \Sexpr{format(round(abs(mean(northamericanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(northamericanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT ($^{\circ}$C) (Fig. \ref{figure:continent}), whereas in Europe these relationships were weaker: advance of \Sexpr{format(round(abs(mean(europeanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(europeanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT $^{\circ}$C(Fig. \ref{figure:continent}). \\

\begin{figure}[!h] 
    \centering
 \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/continent_histogram.png}
    \caption{Caption placeholder}
    \label{figure:continent}
\end{figure}
%EMWApr20 -- please fix all figure edits and add captions for next draft so we can finalize all the tweaks needed for submitting to journal

%EMWApr20 -- add the stats in below paragraph also. Annoying I know, but we're almost done!
Effects of provenance latitude on both spring and fall events were similar across angiosperms and gymnosperms. However, effects of MAT on spring events weakly diverged: spring events shifted earlier as MAT increases in angiosperms and delay as MAT increases in gymnosperms, except for \emph{Pseudotsuga menziesii}. Fall events delay in warmer locations for both species types, but slightly more so for gymnosperms (\Sexpr{format(round(mean(deciduouspost_fitC_fall_mat),digits = 2),nsmall=2)} days VS. \Sexpr{format(round(mean(evergreenspost_fitC_fall_mat),digits = 2),nsmall=2)} days)(Fig. \ref{figure:spp_type}).\\

@@

%EMWApr20 -- need to refer to some supp figures or tables or such below. Once you have the supp so I can see better what we're referring to here in first sentence we should likely edit this paragraph a little. 
We also tested whether climate effects on events were stronger when using metrics from seasonal windows closer to event dates, but results were not qualitatively different than using MAT (See supp? Not sure what results we mean here -- overlap?). We observed very weak effects of climate overlap on spring events, nearly identical across angiosperms and gymnosperms. Fall events diverge as climate overlap declines for both angiosperms and gymnosperms, but slightly more strongly for gymnosperms (Fig.\ref{figure:overlap}).

@@

%EMWApr20 -- update figures with a) Spring events and continent ... b) Fall etc. so they look like ms figures. You also need to add the correct x axis -- they are likely days (spring or fall) per degree latitude or per degree MAT (C) ...
\begin{figure}[!h] 
    \centering
 \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/climate_overlap.png}
    \caption{Caption placeholder}
    \label{figure:overlap}
\end{figure}
%emw6Mar -- Climate overlap figures can go in supp.
